A model including an interaction of treatment with baseline risk was the optimal risk-based approach to predicting absolute treatment benefit

### 2021-07-13 submission/manuscript.rmd
comments on discussion

---
title: "Models including an interaction of treatment with baseline risk were better able to predict absolute benefit in simulations"
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized treatment effects in RCT simulations. **Study Design and Setting:** We simulated data using diverse assumptions for a baseline prognostic index of risk (PI) and the shape of its intereraction with treatment (none, linear or quadratic). In each sample we predicted absolute benefit using: models with the PI and a constant relative treatment effect; models including an interaction of treatment with the PI; stratification in quarters of the PI; nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach using Akaike’s Information Criterion. We evaluated predictive performance using root mean squared error and measures of discrimination and calibration for benefit. Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect strength, and the PI's discriminative ability. **Results:** Models including a PI by treatment interaction performed better under most simulation settings. Flexible models required larger sample sizes and higher AUC of the PI to outperform the linear interaction model. The adaptive approach performed similarly to the best-performing method in most scenarios. **Conclusion:** Usually, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

\linenumbers
# Introduction
Predictive approaches for assessing heterogeneity of treatment effects (HTE) aim
at the development of models predicting either individualized effects or which
of two (or more) treatments is better for an individual [@Varadhan2013]. In
prior work, we divided such methods in three broader categories based on the
reference class used for defining patient similarity when making individualized
predictions or recommendations [@Rekkas2020]. Risk-modeling approaches use
prediction of baseline risk as the reference; treatment effect modeling
approaches also model treatment-covariate interactions, in addition to risk
factors; optimal treatment regime approaches focus on developing treatment
assignment rules and therefore rely heavily on modeling treatment effect
modifiers.

@@

In conclusion, when comparing different risk-based approaches to predict
individualizing benefit a models including a linear treatment interaction with
the prognostic performed best in a wide range of scenarios. More flexible models
with restricted cubic splines required larger sample sizes and higher AUC of the
PI to outperform the linear model. An adaptive approach, selecting the model
with optimal AIC, had comparable performance to the best performing approach in
most of the scenarios.

\newpage
# References
\nolinenumbers
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent
<div id="refs"></div>
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent

>>>>

---
title: |
 A model including an interaction of treatment with baseline risk was the optimal
 risk-based approach to predicting absolute treatment benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its intereraction with treatment (none, linear or quadratic). In 
 each sample we predicted absolute benefit using: models with the PI and a 
 constant relative treatment effect; models including an interaction of treatment 
 with the PI; stratification in quarters of the PI; nonlinear transformations of 
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach 
 using Akaike’s Information Criterion. We evaluated predictive performance using 
 root mean squared error and measures of discrimination and calibration for 
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds 
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect 
 strength, and the PI's discriminative ability. **Results:** Models including a 
 PI by treatment interaction performed better under most simulation settings. 
 Flexible models required larger sample sizes and higher AUC of the PI to 
 outperform the linear interaction model. The adaptive approach performed 
 similarly to the best-performing method in most scenarios. **Conclusion:** 
 Under most circumstances, a model with a linear interaction of the PI with 
 treatment is the optimal risk-based approach to predicting absolute treatment
 benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

# Introduction
Predictive approaches for assessing heterogeneity of treatment effects (HTE) aim
at the development of models predicting either individualized effects or which
of two (or more) treatments is better for an individual [@Varadhan2013]. In
prior work, we divided such methods in three broader categories based on the
reference class used for defining patient similarity when making individualized
predictions or recommendations [@Rekkas2020]. Risk-modeling approaches use
prediction of baseline risk as the reference; treatment effect modeling
approaches also model treatment-covariate interactions, in addition to risk
factors; optimal treatment regime approaches focus on developing treatment
assignment rules and therefore rely heavily on modeling treatment effect
modifiers.

@@

In conclusion, when comparing different risk-based approaches to predicting 
individualized treatment benefit, a model including a linear treatment 
interaction with the prognostic index performed best in a wide range of 
scenarios. More flexible models with restricted cubic splines required larger
sample sizes and higher AUC of the prognostic index to outperform the linear
interaction model. An adaptive approach, selecting the model with the optimal 
AIC, had comparable performance to the best performing approach in most of the 
scenarios.

\newpage
# References
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent
<div id="refs"></div>
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent

### 2021-07-13 submission/manuscript.tex
comments on discussion

\title{Models including an interaction of treatment with baseline risk
were better able to predict absolute benefit in simulations}

@@

\begin{abstract}
\textbf{Objective:} To compare different risk-based methods predicting
individualized treatment effects in RCT simulations. \textbf{Study
Design and Setting:} We simulated data using diverse assumptions for a
baseline prognostic index of risk (PI) and the shape of its
intereraction with treatment (none, linear or quadratic). In each sample
we predicted absolute benefit using: models with the PI and a constant
relative treatment effect; models including an interaction of treatment
with the PI; stratification in quarters of the PI; nonlinear
transformations of the PI (restricted cubic splines with 3, 4 and 5
knots); an adaptive approach using Akaike's Information Criterion. We
evaluated predictive performance using root mean squared error and
measures of discrimination and calibration for benefit. Starting from a
base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
the PI 0.75), we varied the sample size, the treatment effect strength,
and the PI's discriminative ability. \textbf{Results:} Models including
a PI by treatment interaction performed better under most simulation
settings. Flexible models required larger sample sizes and higher AUC of
the PI to outperform the linear interaction model. The adaptive approach
performed similarly to the best-performing method in most scenarios.
\textbf{Conclusion:} Usually, a model with a linear interaction of the
PI with treatment is better able to predict absolute treatment benefit.
\end{abstract}

@@

\linenumbers

@@

In conclusion, when comparing different risk-based approaches to predict
individualizing benefit a models including a linear treatment
interaction with the prognostic performed best in a wide range of
scenarios. More flexible models with restricted cubic splines required
larger sample sizes and higher AUC of the PI to outperform the linear
model. An adaptive approach, selecting the model with optimal AIC, had
comparable performance to the best performing approach in most of the
scenarios.

@@

\nolinenumbers
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}

>>>>

\title{A model including an interaction of treatment with baseline risk
was the optimal risk-based approach to predicting absolute treatment
benefit}

@@

\begin{abstract}
\textbf{Objective:} To compare different risk-based methods predicting
individualized treatment effects in RCT simulations. \textbf{Study
Design and Setting:} We simulated data using diverse assumptions for a
baseline prognostic index of risk (PI) and the shape of its
intereraction with treatment (none, linear or quadratic). In each sample
we predicted absolute benefit using: models with the PI and a constant
relative treatment effect; models including an interaction of treatment
with the PI; stratification in quarters of the PI; nonlinear
transformations of the PI (restricted cubic splines with 3, 4 and 5
knots); an adaptive approach using Akaike's Information Criterion. We
evaluated predictive performance using root mean squared error and
measures of discrimination and calibration for benefit. Starting from a
base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
the PI 0.75), we varied the sample size, the treatment effect strength,
and the PI's discriminative ability. \textbf{Results:} Models including
a PI by treatment interaction performed better under most simulation
settings. Flexible models required larger sample sizes and higher AUC of
the PI to outperform the linear interaction model. The adaptive approach
performed similarly to the best-performing method in most scenarios.
\textbf{Conclusion:} Under most circumstances, a model with a linear
interaction of the PI with treatment is the optimal risk-based approach
to predicting absolute treatment benefit.
\end{abstract}

@@

@@

In conclusion, when comparing different risk-based approaches to
predicting individualized treatment benefit, a model including a linear
treatment interaction with the prognostic index performed best in a wide
range of scenarios. More flexible models with restricted cubic splines
required larger sample sizes and higher AUC of the prognostic index to
outperform the linear interaction model. An adaptive approach, selecting
the model with the optimal AIC, had comparable performance to the best
performing approach in most of the scenarios.

@@

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
